---
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Knitting options

library(dplyr)
library(pander)                           # Load required packages

```

<br>

# Source Data

The analysis uses data comprised of traffic accidents in the city of 
Tempe, Arizona
[**Tempe Open Data Portal**](https://data.tempe.gov/datasets/tempegov::1-08-crash-data-report-detail/about). 

```{r include = FALSE}

# READ IN DATA

url <- paste0("https://github.com/DS4PS/Data-Science-Class/blob",
              "/master/DATA/TempeTrafficAccidents.rds?raw=true")

dat <- readRDS(gzcon(url(url)))     # Method per instructions

```

<br>

## Preview

```{r}

head(dat)

```
<br>

# Preprocessing

```{r eval = FALSE, echo = TRUE, include = TRUE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     levels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>

```{r include = FALSE}

date.vec <- strptime(dat$DateTime, 
                     format = "%m/%d/%y %H:%M")       # Create date fields

dat$hour   <- format(date.vec, format = "%H")
dat$month  <- format(date.vec, format = "%b")
dat$day    <- format(date.vec, format = "%a")
dat$day365 <- format(date.vec, format = "%j")
dat$week   <- format(date.vec, format = "%V")

dat$day <- factor(dat$day, 
                  levels = c("Mon",
                             "Tue",
                             "Wed",
                             "Thu",
                             "Fri",
                             "Sat",
                             "Sun"))                  # Order weekdays

dat$hour12 <- format(date.vec, 
                     format="%l %p")                  # Create 12-hour format

time.levels <- c("12 AM", paste(1:11, "AM"), 
                 "12 PM", paste(1:11, "PM"))

dat$hour12 <- factor(dat$hour12, 
                     labels = time.levels)            # Order time intervals

age.labels <- paste0("Age ", 
                     c(16,18,25,35,45,55,65,75), "-", 
                     c(18,25,35,45,55,65,75,100) )

dat$age <- cut(dat$Age_Drv1, 
               breaks = c(16,18,25,
                          35,45,55,
                          65,75,100), 
               labels = age.labels)                   # Discretize age ranges

```

<br>

# Part I: Summaries

## 1. Mondays Accidents
<br>

```{r}

sum(dat$day == "Mon", 
    na.rm = TRUE)

```

<br>

## 2. Monday Accidents (%)

Calculate the proportion of accidents occur on Monday each week.

<br>

```{r}

proportion_monday <- dat %>%
  filter(day == "Mon") %>%
  summarise(proportion = n() / nrow(dat))

```

<br>

**Answer:** **0.1438005%**

<br>

## 3. Harmful Monday Accidents (%)
The proportion of accidents on mondays result in harm

<br>

```{r}
proportion_monday_harm <- dat %>%
  filter(day == "Mon" & (Totalinjuries > 0 | Totalfatalities > 0)) %>%
  summarise(proportion = n() / nrow(dat))

```

<br>

**Answer:** **0.04292237%**

<br>

## 4. Most Common Accidents
The most common accident type that occurs on monday.

<br>

```{r}

dat %>%
  filter(day == "Mon") %>%
  count(Collisionmanner) %>%
  arrange(desc(n))

```

<br>

**Answer:** **Rear End**

<br>

## 5. Differences in Accidents
The difference in the proportion of accidents resulting in harm across the days of the week.

<br>

```{r}

proportion_harm_by_day <- dat %>%
  group_by(day) %>%
  summarise(proportion_harm = sum(Totalinjuries > 0 | Totalfatalities > 0) / n())

proportion_harm_by_day

```

<br>

**Answer:** More accidents occur **Thursday**.

<br>

## 6. Accident Stats by Weekday
<br>

```{r}

dat %>%
  group_by(day) %>%
  mutate(harm_acc = Totalinjuries > 0 | Totalfatalities > 0) %>%
  summarize(n = n(),
            injuries = sum(Totalinjuries),
            fatalities = sum(Totalfatalities),
            harm.rate = mean(harm_acc))

```

<br>
<br>

# Part II: Age Groups


## 1. Accidents by Hour & Age
The age group that had the largest number of accident at 7Am.

<br>

```{r}

accident_counts <- dat %>%
  count(hour12, age)

accident_counts

accidents_7am <- accident_counts %>%
  filter(hour12 == "7 AM")

largest_age_group_7am <- accidents_7am %>%
  filter(n == max(n)) %>%
  pull(age)
```

<br>

**Answer:** **18-25**

<br>

## 2. Accidents by Hour & Age (%)

```{r echo = FALSE}
data.frame(age = c("Age 16-18", "Age 18-25", "Age 25-35"),
           hour12 = c("7 AM", "7 AM", "7 AM"),
           n = c(77, 408, 371),
           n.age = c(1458, 8796, 5456),
           n.hour = c(1606, 1606, 1606),
           p = c(0.05, 0.25, 0.23),
           p.age = c(0.05, 0.05, 0.07),
           p.hour = c(0.05, 0.25, 0.23)) %>%
  pander()


accidents_7am <- accident_counts %>%
  filter(hour12 == "7 AM")

proportions <- accidents_7am %>%
  mutate(n.age = sum(n)) %>%
  group_by(age) %>%
  mutate(p = n / sum(n),
         p.age = sum(n) / sum(n.age),
         p.hour = n / sum(n)) %>%
  ungroup() %>%
  select(age, hour12, n, n.age, p, p.age, p.hour)

proportions

```

<br>

```{r}

dat %>% 
  group_by(hour12, age) %>% 
  summarize(n = n()) %>%      # Partial solution
  group_by(age) %>%           # Requires two new variables in 'mutate()'
  mutate()                    # Lastly, requires function 'filter()

```

<br>
<br>

# Part III: Rates of Harm

## 1. Accidents by Hour

<br>

```{r}

dat %>%
  group_by(hour) %>%
  summarize(n = n()) %>%
  plot(type = "b",
       bty = "n",
       pch = 19,
       cex = 2,
       xlab = "Hour",
       ylab = "Total Number of Accidents",
       main = "Total Number of Accidents by Time of Day")

```

<br>

## 2. Total Casualties by Hour

<br>

```{r}

injuries_fatalities_sum <- dat %>%
  group_by(hour) %>%
  summarise(TotalInjuries = sum(Totalinjuries))

plot(injuries_fatalities_sum$hour, injuries_fatalities_sum$TotalInjuries,
     type = "b",
     bty = "n",
     pch = 19,
     cex = 2,
     xlab = "Hour",
     ylab = "Total Number of passengers hurt or killed",
     main = "Total Injuries by hour of Day")

```

<br>

## 3. Accidents with Casualties (%)

<br>

```{r}

proportion_harmful <- dat %>%
  group_by(hour) %>%
  summarize(ProportionHarmful = sum(Totalinjuries > 0 | Totalfatalities > 0) / n())

# Create a line plot
plot(proportion_harmful$hour, proportion_harmful$ProportionHarmful, 
     type = "b",
     bty = "n",
     pch = 19,
     cex = 2,
     xlab = "Hour",
     ylab = "Proportion of Harmful Accidents",
     main = "Proportion of Accidents result in injuries or fatalities")

```

<br>

## 4. Avg. Harm by Harmful Accident

<br>

```{r}

average_injuries_fatalities <- dat %>%
  filter(Totalinjuries > 0 | Totalfatalities > 0) %>%
  group_by(hour) %>%
  summarize(AvgInjuriesFatalities = mean(Totalinjuries + Totalfatalities))

plot(average_injuries_fatalities$hour, average_injuries_fatalities$AvgInjuriesFatalities,
     type = "b",
     bty = "n",
     pch = 19,
     cex = 2,
     xlab = "Hour",
     ylab = "Average number of passengers hurts",
     main = "Average Injuries or Fatalities per Harmful Accidents")

```
